Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-02-03 Manuscript/manuscript.Rmd
Methods proof-reading

Replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 4) [??? Need supplementary ref.; !!!added supp table] The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained.

@@

Genomic annotation data (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 1). The annotations consisted of a variety of different chromatin states, histone marks, and transcription factor binding sites using ChIP-seq peak experiments for the GM12878, K562, ... cell lines. Each genomic annotation was represented by a set of regions (their genomic coordinates) annotated as having a functional property or biological significance.

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). For example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. Returning to the 10 kb resolution example, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

The data sets used for modeling were composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. These two classes made up the minority and majority sets respectively. The data was then split into a 7:3 ratio of training to testing sets. Each training and testing set was composed of a similar majority to minority class ratio as the full data set. For the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for a fair and unbiased comparison.

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance (refered to hereafter simply as $\textit{distance}$), in basepairs, from the center of the bin to the center of the nearest region defined by a specific genomic annotation [??? Decide whether to use "annotation" or "feature"; !!!go with annotation] was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figures 1-8).

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. The evaluation metrics listed above are all considered single threshold measures. That is, they are defined for individual thresholds that separated class prediction. Thererfore they are unable to give an overview of the potential range of performance as thresholds vary. In order to evaluate the trade-off between precision and recall, we also considered a threshold-free measure in form of the area under the precision recall curve (AUPRC).

###MODEL ASSESSMENT AND PREDICTION

We first assessed predictor type performance using random forest classification algorithms by utilizing each set of predictor types as the feature space, in addition to a model with all combined types. Models were evaluated across each combination of resampling technique applied to the training set, for each cell line. Performance was based on the AUPRC. Recursive feature elimination (RFE) was then used to determine the optimal number of features for models with the best performing predictor type, across resampling techniques and cell lines. In doing so, random forest models were again used to recursively assign weights to a set of features of a given size. For computational efficiency we evaluated feature subset sizes in descending order of powers of 2. The optimal number of features was determined by the AUPRC of the corresponding model. Optimal performance was assessed and the feature spaced was reduced accordingly.

For the task of identifying the functional genomic elements that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each predictor of the random forest models associated with each resampling technique. For each model and within each tree, the AUPRC on the out-of-bag portion of the data was recorded. Then the same was done after permuting each predictor variable. The difference between the two AUPRC values was then averaged over all trees, and normalized by the standard error. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret].

>>>>

Replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 4). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained.

@@

Genomic annotation data (chromatin states, histone marks, and transcription factor binding sites, GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 1).

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). For example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. For example, for 10 kb data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0) (Figure 1).

@@

The data sets used for modeling were composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. These two classes made up the minority and majority sets respectively. The data was then split into a 7:3 ratio of training to testing sets. Each training and testing set was composed of a similar majority to minority class ratio as the full data set. For the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison.

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation [??? Decide whether to use "annotation" or "feature"; !!!go with annotation] was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figures 1-8).

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. The evaluation metrics listed above are all considered single threshold measures. That is, they are defined for individual thresholds that separated class prediction. Thererfore, they are unable to give an overview of the potential range of performance as thresholds vary. In order to evaluate the trade-off between precision and recall, we also considered a threshold-free measure in form of the area under the precision recall curve (AUPRC). [??? This seem to be a good place to describe it in more details]

### MODEL ASSESSMENT AND PREDICTION

We first assessed predictor type performance using random forest classification algorithms by utilizing each set of predictor types as the feature space, in addition to a model with all types combined. Models were evaluated across each combination of resampling technique applied to the training set, for each cell line. Performance was based on the AUPRC. Recursive feature elimination (RFE) was then used to determine the optimal number of features for models with the best performing predictor type, across resampling techniques and cell lines. [??? Unclear at shich step this was done] In doing so, random forest models were again used to recursively assign weights to a set of features of a given size. For computational efficiency we evaluated feature subset sizes in descending order of powers of 2. The optimal number of features was determined by the AUPRC of the corresponding model. Optimal performance was assessed and the feature spaced was reduced accordingly.

For the task of identifying the functional genomic elements that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each predictor of the random forest models associated with each resampling technique. For each model and within each tree, the AUPRC on the out-of-bag portion of the data was recorded. Then the same was done after permuting each predictor variable. The difference between the two AUPRC values was then averaged over all trees, and normalized by the standard error [??? of what?]. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret].
